Artificial intelligence-assisted analysis of CT abnormalities during COVID-19 recovery

### 2025-03-14 paper/markdown/manuscript.Rmd
Tuning grids in Supplementary Material
Showing the complete tuning grids in Supplementary Tables. Supplementary Figures for tuning of the models of DLCO < 80% and DLCO

Prediction of lung function deficits following COVID-19 with demographic, clinical and biochemical explanatory factors and characteristic of acute SARS-CoV-2 infection suffers from low accuracy [@Sonnweber2022]. 
Additionally, relevance of residual structural lung lesions and severity of lung damage in COVD-19 convalescents for pulmonary function is not entirely clear [@Luger2022; @Sahanic2023]. 
Herein, we sought to model the major LFT abnormalities, i.e. insufficient DLCO, FVC, and FEV1, and numeric readouts, i.e DLCO, FVC, FEV1 in the CovILD cohort with a broad panel of `r lft_globals[c("ct_variables", "symptom_variables", "baseline_variables")] %>% reduce(union) %>% length` independent variables including baseline demographic and clinical characteristic (e.g. gender, age, smoking history, comorbidity), characteristic of acute COVID-19 (e.g. severity, medication, hospitalization) and convalescence (follow-up time point, persistent symptoms), as well as presence and severity CT abnormalities (e.g. GGO, CTSS, AI-determined opacity and high opacity, __Supplementary Table S2__). 
To this end, we employed four popular machine learning algorithms, 
Random Forest [@Breiman2001], 
gradient boosted machines (GBM) [@Friedman2002; @Friedman2001], 
neural network with a single hidden layer [@Ripley2014], and 
support vector machines (SVM) with radial kernel [@Weston1998]. 
Because the CovILD data set consisted of non-independent, participant-matched observations, blocked cross-validation with participant's identifier serving as the grouping variable was used for optimization of the model hyper-parameters (tuning, __Supplementary Table S6__). 
Predictions of the models were finally evaluated by comparison with the observed outcome in the training data and the blocked cross-validation folds [@Kuhn2008] (__Supplementary Figure S8__, __Table \@ref(tab:tab-bin-classifiers)__ and __\@ref(tab:tab-reg-models)__,  __Supplementary Tables S7__ and __S8__).

Among the investigated LFT abnormalities, solely insufficient DLCO defined as values < 80% of the patient's reference yielded meaningful models as evaluated by cross-validation; accuracy of the models of DLCO < 80% was reproducible between the algorithms (overall accuracy: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$correct_rate %>% range %>% signif(2) %>% paste(collapse = ' to ')`, Cohen's $\kappa$: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$kappa %>% range %>% signif(2) %>% paste(collapse = ' to ')`, AUC: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$AUC %>% range %>% signif(2) %>% paste(collapse = ' to ')`). 
Low cross-validated Brier score values for the models of insufficient DLCO indicated good overall calibration (Brier score: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$brier_score %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Figure \@ref(fig:fig-ml-performance)A__). 
In turn, models of insufficient FVC and FEV1 performed poorly in the cross-validation setting (overall accuracy: `r bin_models$stats[c("FVC_reduced", "FEV1_reduced")] %>% map(filter, dataset == 'cv') %>% map(~.x$correct_rate) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, Cohen's $\kappa$: `r bin_models$stats[c("FVC_reduced", "FEV1_reduced")] %>% map(filter, dataset == 'cv') %>% map(~.x$kappa) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, AUC: `r bin_models$stats[c("FVC_reduced", "FEV1_reduced")] %>% map(filter, dataset == 'cv') %>% map(~.x$AUC) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Supplementary Figure S9__).

As in case of the LFT abnormalities, solely DLCO expressed as percentages of the patient's reference was modeled with low error and substantial explanatory power in cross-validation with the Random Forest, GBM, and SVM algorithms (MAE: `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm != 'nnet') %>%  .$MAE %>% range %>% signif(3) %>% paste(collapse = ' to ')`, pseudo-R^2^: `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm != 'nnet') %>%  .$rsq %>% range %>% signif(2) %>% paste(collapse = ' to ')`). 
Those models were also characterized by good calibration assessed by Spearman's $\rho$ coefficients of correlation between the predicted and observed DLCO (Spearman's $\rho$: `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm != 'nnet') %>%  .$spearman %>% range %>% signif(2) %>% paste(collapse = ' to ')`). 
By contrast, the neural network model of DLCO suffered from large error and low fraction of explained variance in the cross-validation setting (MAE = `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm == 'nnet') %>%  .$MAE %>% signif(3)`, pseudo-R^2^ = `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm == 'nnet') %>%  .$rsq %>% signif(2)`, __Figure \@ref(fig:fig-ml-performance)B__). 
No meaningful models of FVC or FEV1 could be developed with any of the machine learning algorithms (cross-validation, R^2^: `r reg_models$stats[c("FVC_percent", "FEV1_percent")] %>% map(filter, dataset == 'cv') %>% map(~.x$rsq) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Supplementary Figure S10__).

Concerning the models of insufficient DLCO, the highest concordance between the predicted and observed DLCO < 80% of the patient's reference was achieved for moderate COVID-19 convalescents at the two- to six-month follow-up examinations (all algorithms, $\kappa$: `r dlco_resid$stats$kappa %>% map(filter, severity_class == 'hospitalized moderate', follow_up %in% c('2-month follow-up', '3-month follow-up', '6-month follow-up')) %>% map(~.x$kappa) %>% reduce(c) %>% range %>% signif(2)`). 
The accuracy was in turn the poorest for ambulatory patients at the six- and twelve-month follow-up (all algorithms, $\kappa$: `r dlco_resid$stats$kappa %>% map(filter, severity_class == 'ambulatory mild', follow_up %in% c('6-month follow-up', '12-month follow-up')) %>% map(~.x$kappa) %>% reduce(c) %>% range %>% signif(2)`, __Supplementary Figure S11A__). 
The Random Forest, SVM and GBM models of DLCO expressed as percentage of the patient's reference exhibited the lowest average errors for moderate COVID-19 survivors (mean error: `r dlco_resid$stats$DLCO_residuals[c("ranger", "svmRadial", "gbm")] %>% map(filter, severity_class == 'hospitalized moderate') %>% map(~.x$mean) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`). 
In turn, DLCO values were systematically overestimated for severe COVID-19 patients at the two- to six-month follow-up visits (mean error: `r dlco_resid$stats$DLCO_residuals[c("ranger", "svmRadial", "gbm")] %>% map(filter, severity_class == 'hospitalized severe', follow_up %in% c('2-month follow-up', '3-month follow-up', '6-month follow-up')) %>% map(~.x$mean) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`) and underestimated for ambulatory COVID-19 convalescents at the same time points (mean error: `r dlco_resid$stats$DLCO_residuals[c("ranger", "svmRadial", "gbm")] %>% map(filter, severity_class == 'ambulatory mild', follow_up %in% c('2-month follow-up', '3-month follow-up', '6-month follow-up')) %>% map(~.x$mean) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Supplementary Figure S11B__). 
The best performance in the hospitalized moderate subset is likely attributed to the large number of observations and frequent DLCO deficits in this group of participants available for training of the models (n = `r lft_globals$analysis_tbl %>% filter(severity_class == 'hospitalized moderate', DLCO_reduced == 'yes') %>% nrow` observations of DLCO < 80% in `r lft_globals$analysis_tbl %>% filter(severity_class == 'hospitalized moderate') %>% nrow` observations from moderate COVID-19 patients). 
Conversely, the numbers of observations from ambulatory COVID-19 patients and numbers of observations DLCO insufficiency in this group was low (`r lft_globals$analysis_tbl %>% filter(severity_class == 'ambulatory mild', DLCO_reduced == 'yes') %>% nrow` observations of DLCO < 80% in `r lft_globals$analysis_tbl %>% filter(severity_class == 'ambulatory mild') %>% nrow` observations from ambulatory patients) which resulted in relatively poor performance of the machine learning models.

Prediction of reduced DLCO < 80% was assessed in observations with increasing values of DLCO expressed as percentage of the patient's reference by moving averages of accuracy and error metered as squared distance to the 0/1-coded outcome. 
As presented in __Supplementary Figure S12__ for the best performing GBM model, accuracy was the lowest and the error peaked for observations with DLCO ranging between 70% and 80% of the patient's reference. 
An analogical phenomenon was observed for the remaining machine learning algorithms (not shown). 
This illustrates that while predictions of highly compromised and normal-to-high DLCO are accurate, forecasts for borderline cases of reduced DLCO suffer from high error. 
This may question biological and clinical relevance of the 80% cutoff widely used for diagnosis of insufficient DLCO in COVID-19 survivors [@Sonnweber2020; @Zhang2022a].

To assess under- and over-fitting, we resorted to analyses of learning curves of the models of DLCO and reduced DLCO. With this approach, we evaluated performance of the models re-trained in subsets of the modeling data set of varying sizes. 
The performance evaluation was done for the training subsets, test subsets (one-forth of observations not used for the model training), and 10-repeats 10-fold cross-validation [@Viering2023]. 
The learning curves for the best-performing GBM models are presented in __Supplementary Figure S13__, for the remaining algorithms comparable results were obtained (not shown). 
As inferred from substantial differences of accuracy and Cohen's $\kappa$ between the training, test, and cross-validation subsets even for the largest sizes of the training data, models of insufficient DLCO suffer from over-fitting, i.e. poor generalizability for unseen data. 
The generalizability of the models of DLCO expressed as percentage of the patient's reference was substantially better.

As investigated by absolute values of SHAP variable importance metrics [@Covert2020; @Lundberg2017], human- and AI-derived ratings of structural lung damage, CTSS, opacity and high opacity, belonged to the most influential explanatory variables of the models of DLCO insufficiency and the meaningful models of DLCO. 
Other influential explanatory variables were 
well characterized risk factors of severe COVID-19 (age, male sex, body mass index, pre-existing malignancy and cardiovascular disease), 
readouts of acute COVID-19 severity (severity class, WHO ordinal severity scale, hospitalization length, ICU treatment, anti-coagulant treatment, and weight change), 
smoking intensity (pack-years), as well as 
rating of physical performance impairment (ECOG) and 
exertional dyspnea (mMRC) (__Figure \@ref(fig:fig-varimp-dlco-red)__, __Supplementary Figure S14__, __Supplementary Table S9__).

Analysis of a correlation graph of the most influential explanatory factors for the predictions of DLCO and insufficient DLCO revealed associations between CT readouts of lung damage severity (CTSS, opacity, and high opacity, presence of any lesions, GGO and reticulations) and between readouts of acute COVID-19 severity (COVID-19 severity, ICU stay, hospital stay, anti-infective treatment during acute infection) (__Supplementary Figure S15A__). 
In particular, CTSS and opacity (correlation coefficient, Spearman's $\rho$ = `r corr_lft$test$ct %>% filter(variable1 == 'CTSS', variable2 == 'opacity_percent') %>% .$rho %>% signif(2)`), CTSS and high opacity ($\rho$ = `r corr_lft$test$ct %>% filter(variable1 == 'CTSS', variable2 == 'high_opacity_percent') %>% .$rho %>% signif(2)`), and opacity and high opacity (`r corr_lft$test$ct %>% filter(variable1 == 'opacity_percent', variable2 == 'high_opacity_percent') %>% .$rho %>% signif(2)`) were strongly inter-correlated. 
CTSS values were significantly higher in observations with than without 
GGO (difference of medians: `r blast(comp_ct$analysis_tbl, GGO_finding) %>% map(~.x$CTSS) %>% map_dbl(median) %>% diff` points, `r comp_ct$test %>% filter(variable == 'CTSS', split_factor == 'GGO_finding') %>% .$significance`, effect size: `r comp_ct$test %>% filter(variable == 'CTSS', split_factor == 'GGO_finding') %>% .$eff_size`) and 
reticulation (difference of medians: `r blast(comp_ct$analysis_tbl, reticulation_finding) %>% map(~.x$CTSS) %>% map_dbl(median) %>% diff` points, `r comp_ct$test %>% filter(variable == 'CTSS', split_factor == 'reticulation_finding') %>% .$significance`, effect size: `r comp_ct$test %>% filter(variable == 'CTSS', split_factor == 'reticulation_finding') %>% .$eff_size`) (__Supplementary Figure S3B__, __Supplementary Table S10__). 
Analogically, significantly higher opacity was detected in observations with than without 
GGO (difference of medians: `r blast(comp_ct$analysis_tbl, GGO_finding) %>% map(~.x$opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`% of lung, `r comp_ct$test %>% filter(variable == 'opacity_percent', split_factor == 'GGO_finding') %>% .$significance`, effect size: `r comp_ct$test %>% filter(variable == 'opacity_percent', split_factor == 'GGO_finding') %>% .$eff_size`) and 
reticulation (difference of medians: `r blast(comp_ct$analysis_tbl, reticulation_finding) %>% map(~.x$opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`% of lung, `r comp_ct$test %>% filter(variable == 'opacity_percent', split_factor == 'reticulation_finding') %>% .$significance`, effect size: `r comp_ct$test %>% filter(variable == 'opacity_percent', split_factor == 'reticulation_finding') %>% .$eff_size`). 
Effect sizes of these differences were large (__Supplementary Figure S15B__, __Supplementary Tables S11__).
Collectively, the high co-linearity of the most important explanatory variables points suggests their redundancy. 
Tackling this redundancy e.g. by regularization may control the over-fitting and improve accuracy of predictions of DLCO and reduced DLCO by machine learning algorithms.

@@

Human-determined CTSS (difference of medians: `r blast(lft_uniboot$analysis_tbl, DLCO_reduced) %>% map(~.x$CTSS) %>% map_dbl(median) %>% diff` points, `r filter(lft_uniboot$test, split_factor == 'DLCO_reduced', variable == 'CTSS')$significance`, effect size: `r filter(lft_uniboot$test, split_factor == 'DLCO_reduced', variable == 'CTSS')$eff_size`), 
AI-determined opacity (difference of medians: `r blast(lft_uniboot$analysis_tbl, DLCO_reduced) %>% map(~.x$opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`% of the lung, `r filter(lft_uniboot$test, split_factor == 'DLCO_reduced', variable == 'opacity_percent')$significance`, effect size: `r filter(lft_uniboot$test, split_factor == 'DLCO_reduced', variable == 'opacity_percent')$eff_size`) and 
high opacity (difference of medians: `r blast(lft_uniboot$analysis_tbl, DLCO_reduced) %>% map(~.x$high_opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`, `r filter(lft_uniboot$test, split_factor == 'DLCO_reduced', variable == 'high_opacity_percent')$significance`% of the lung, effect size: `r filter(lft_uniboot$test, split_factor == 'DLCO_reduced', variable == 'high_opacity_percent')$eff_size`) 
were significantly higher in observations with DLCO < 80% of the patient's reference than in the remaining data points. 
The effect size of those differences was large (__Figure \@ref(fig:fig-dlco-red-detection)A__). 
AI-measured opacity of the lung, but not CTSS or high opacity, was found to be significantly higher in observations with insufficient 
FVC (difference of medians: `r blast(lft_uniboot$analysis_tbl, FVC_reduced) %>% map(~.x$opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`, `r filter(lft_uniboot$test, split_factor == 'FVC_reduced', variable == 'opacity_percent')$significance`, effect size: `r filter(lft_uniboot$test, split_factor == 'FVC_reduced', variable == 'opacity_percent')$eff_size`) and 
insufficient FEV1 (difference of medians: `r blast(lft_uniboot$analysis_tbl, FEV1_reduced) %>% map(~.x$opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`, `r filter(lft_uniboot$test, split_factor == 'FEV1_reduced', variable == 'opacity_percent')$significance`, effect size: `r filter(lft_uniboot$test, split_factor == 'FEV1_reduced', variable == 'opacity_percent')$eff_size`); sizes of those effects were, however, small (__Supplementary Table S12__). 
Furthermore, CTSS, opacity and high opacity correlated negatively with moderate effect size with DLCO expressed as percentage of the patient's reference (p < 0.001, Spearman's $\rho$: `r lft_uniboot$cor_test %>% filter(variable1 == 'DLCO_percent') %>% .$rho %>% range %>% signif(2) %>% paste(collapse = ' to ')`). 
Interestingly, the CT readouts of structural lung damage also correlated significantly with FVC (Spearman's $\rho$: `r lft_uniboot$cor_test %>% filter(variable1 == 'FVC_percent') %>% .$rho %>% range %>% signif(2) %>% paste(collapse = ' to ')`) and FEV1 (Spearman's $\rho$: `r lft_uniboot$cor_test %>% filter(variable1 == 'FEV1_percent') %>% .$rho %>% range %>% signif(2) %>% paste(collapse = ' to ')`); effect size of those associations was small (__Supplementary Table S13__).

>>>>

Prediction of lung function deficits following COVID-19 with demographic, clinical and biochemical explanatory factors and characteristic of acute SARS-CoV-2 infection suffers from low accuracy [@Sonnweber2022]. 
Additionally, relevance of residual structural lung lesions and severity of lung damage in COVD-19 convalescents for pulmonary function is not entirely clear [@Luger2022; @Sahanic2023]. 
Herein, we sought to model the major LFT abnormalities, i.e. insufficient DLCO, FVC, and FEV1, and numeric readouts, i.e DLCO, FVC, FEV1 in the CovILD cohort with a broad panel of `r lft_globals[c("ct_variables", "symptom_variables", "baseline_variables")] %>% reduce(union) %>% length` independent variables including baseline demographic and clinical characteristic (e.g. gender, age, smoking history, comorbidity), characteristic of acute COVID-19 (e.g. severity, medication, hospitalization) and convalescence (follow-up time point, persistent symptoms), as well as presence and severity CT abnormalities (e.g. GGO, CTSS, AI-determined opacity and high opacity, __Supplementary Table S2__). 
To this end, we employed four popular machine learning algorithms, 
Random Forest [@Breiman2001], 
gradient boosted machines (GBM) [@Friedman2002; @Friedman2001], 
neural network with a single hidden layer [@Ripley2014], and 
support vector machines (SVM) with radial kernel [@Weston1998]. 
Because the CovILD data set consisted of non-independent, participant-matched observations, blocked cross-validation with participant's identifier serving as the grouping variable was used for optimization of the model hyper-parameters (tuning, __Supplementary Tables S6__ and __S7__, __Supplementary Figures S8__ - __S10__). 
Predictions of the models were finally evaluated by comparison with the observed outcome in the training data and the blocked cross-validation folds [@Kuhn2008] (__Supplementary Figure S8__, __Table \@ref(tab:tab-bin-classifiers)__ and __\@ref(tab:tab-reg-models)__,  __Supplementary Tables S8__ and __S9__).

Among the investigated LFT abnormalities, solely insufficient DLCO defined as values < 80% of the patient's reference yielded meaningful models as evaluated by cross-validation; accuracy of the models of DLCO < 80% was reproducible between the algorithms (overall accuracy: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$correct_rate %>% range %>% signif(2) %>% paste(collapse = ' to ')`, Cohen's $\kappa$: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$kappa %>% range %>% signif(2) %>% paste(collapse = ' to ')`, AUC: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$AUC %>% range %>% signif(2) %>% paste(collapse = ' to ')`). 
Low cross-validated Brier score values for the models of insufficient DLCO indicated good overall calibration (Brier score: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$brier_score %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Figure \@ref(fig:fig-ml-performance)A__). 
In turn, models of insufficient FVC and FEV1 performed poorly in the cross-validation setting (overall accuracy: `r bin_models$stats[c("FVC_reduced", "FEV1_reduced")] %>% map(filter, dataset == 'cv') %>% map(~.x$correct_rate) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, Cohen's $\kappa$: `r bin_models$stats[c("FVC_reduced", "FEV1_reduced")] %>% map(filter, dataset == 'cv') %>% map(~.x$kappa) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, AUC: `r bin_models$stats[c("FVC_reduced", "FEV1_reduced")] %>% map(filter, dataset == 'cv') %>% map(~.x$AUC) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Supplementary Figure S11__).

As in case of the LFT abnormalities, solely DLCO expressed as percentages of the patient's reference was modeled with low error and substantial explanatory power in cross-validation with the Random Forest, GBM, and SVM algorithms (MAE: `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm != 'nnet') %>%  .$MAE %>% range %>% signif(3) %>% paste(collapse = ' to ')`, pseudo-R^2^: `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm != 'nnet') %>%  .$rsq %>% range %>% signif(2) %>% paste(collapse = ' to ')`). 
Those models were also characterized by good calibration assessed by Spearman's $\rho$ coefficients of correlation between the predicted and observed DLCO (Spearman's $\rho$: `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm != 'nnet') %>%  .$spearman %>% range %>% signif(2) %>% paste(collapse = ' to ')`). 
By contrast, the neural network model of DLCO suffered from large error and low fraction of explained variance in the cross-validation setting (MAE = `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm == 'nnet') %>%  .$MAE %>% signif(3)`, pseudo-R^2^ = `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm == 'nnet') %>%  .$rsq %>% signif(2)`, __Figure \@ref(fig:fig-ml-performance)B__). 
No meaningful models of FVC or FEV1 could be developed with any of the machine learning algorithms (cross-validation, R^2^: `r reg_models$stats[c("FVC_percent", "FEV1_percent")] %>% map(filter, dataset == 'cv') %>% map(~.x$rsq) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Supplementary Figure S12__).

Concerning the models of insufficient DLCO, the highest concordance between the predicted and observed DLCO < 80% of the patient's reference was achieved for moderate COVID-19 convalescents at the two- to six-month follow-up examinations (all algorithms, $\kappa$: `r dlco_resid$stats$kappa %>% map(filter, severity_class == 'hospitalized moderate', follow_up %in% c('2-month follow-up', '3-month follow-up', '6-month follow-up')) %>% map(~.x$kappa) %>% reduce(c) %>% range %>% signif(2)`). 
The accuracy was in turn the poorest for ambulatory patients at the six- and twelve-month follow-up (all algorithms, $\kappa$: `r dlco_resid$stats$kappa %>% map(filter, severity_class == 'ambulatory mild', follow_up %in% c('6-month follow-up', '12-month follow-up')) %>% map(~.x$kappa) %>% reduce(c) %>% range %>% signif(2)`, __Supplementary Figure S13A__). 
The Random Forest, SVM and GBM models of DLCO expressed as percentage of the patient's reference exhibited the lowest average errors for moderate COVID-19 survivors (mean error: `r dlco_resid$stats$DLCO_residuals[c("ranger", "svmRadial", "gbm")] %>% map(filter, severity_class == 'hospitalized moderate') %>% map(~.x$mean) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`). 
In turn, DLCO values were systematically overestimated for severe COVID-19 patients at the two- to six-month follow-up visits (mean error: `r dlco_resid$stats$DLCO_residuals[c("ranger", "svmRadial", "gbm")] %>% map(filter, severity_class == 'hospitalized severe', follow_up %in% c('2-month follow-up', '3-month follow-up', '6-month follow-up')) %>% map(~.x$mean) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`) and underestimated for ambulatory COVID-19 convalescents at the same time points (mean error: `r dlco_resid$stats$DLCO_residuals[c("ranger", "svmRadial", "gbm")] %>% map(filter, severity_class == 'ambulatory mild', follow_up %in% c('2-month follow-up', '3-month follow-up', '6-month follow-up')) %>% map(~.x$mean) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Supplementary Figure S13B__). 
The best performance in the hospitalized moderate subset is likely attributed to the large number of observations and frequent DLCO deficits in this group of participants available for training of the models (n = `r lft_globals$analysis_tbl %>% filter(severity_class == 'hospitalized moderate', DLCO_reduced == 'yes') %>% nrow` observations of DLCO < 80% in `r lft_globals$analysis_tbl %>% filter(severity_class == 'hospitalized moderate') %>% nrow` observations from moderate COVID-19 patients). 
Conversely, the numbers of observations from ambulatory COVID-19 patients and numbers of observations DLCO insufficiency in this group was low (`r lft_globals$analysis_tbl %>% filter(severity_class == 'ambulatory mild', DLCO_reduced == 'yes') %>% nrow` observations of DLCO < 80% in `r lft_globals$analysis_tbl %>% filter(severity_class == 'ambulatory mild') %>% nrow` observations from ambulatory patients) which resulted in relatively poor performance of the machine learning models.

Prediction of reduced DLCO < 80% was assessed in observations with increasing values of DLCO expressed as percentage of the patient's reference by moving averages of accuracy and error metered as squared distance to the 0/1-coded outcome. 
As presented in __Supplementary Figure S14__ for the best performing GBM model, accuracy was the lowest and the error peaked for observations with DLCO ranging between 70% and 80% of the patient's reference. 
An analogical phenomenon was observed for the remaining machine learning algorithms (not shown). 
This illustrates that while predictions of highly compromised and normal-to-high DLCO are accurate, forecasts for borderline cases of reduced DLCO suffer from high error. 
This may question biological and clinical relevance of the 80% cutoff widely used for diagnosis of insufficient DLCO in COVID-19 survivors [@Sonnweber2020; @Zhang2022a].

To assess under- and over-fitting, we resorted to analyses of learning curves of the models of DLCO and reduced DLCO. With this approach, we evaluated performance of the models re-trained in subsets of the modeling data set of varying sizes. 
The performance evaluation was done for the training subsets, test subsets (one-forth of observations not used for the model training), and 10-repeats 10-fold cross-validation [@Viering2023]. 
The learning curves for the best-performing GBM models are presented in __Supplementary Figure S15__, for the remaining algorithms comparable results were obtained (not shown). 
As inferred from substantial differences of accuracy and Cohen's $\kappa$ between the training, test, and cross-validation subsets even for the largest sizes of the training data, models of insufficient DLCO suffer from over-fitting, i.e. poor generalizability for unseen data. 
The generalizability of the models of DLCO expressed as percentage of the patient's reference was substantially better.

As investigated by absolute values of SHAP variable importance metrics [@Covert2020; @Lundberg2017], human- and AI-derived ratings of structural lung damage, CTSS, opacity and high opacity, belonged to the most influential explanatory variables of the models of DLCO insufficiency and the meaningful models of DLCO. 
Other influential explanatory variables were 
well characterized risk factors of severe COVID-19 (age, male sex, body mass index, pre-existing malignancy and cardiovascular disease), 
readouts of acute COVID-19 severity (severity class, WHO ordinal severity scale, hospitalization length, ICU treatment, anti-coagulant treatment, and weight change), 
smoking intensity (pack-years), as well as 
rating of physical performance impairment (ECOG) and 
exertional dyspnea (mMRC) (__Figure \@ref(fig:fig-varimp-dlco-red)__, __Supplementary Figure S16__, __Supplementary Table S10__).

Analysis of a correlation graph of the most influential explanatory factors for the predictions of DLCO and insufficient DLCO revealed associations between CT readouts of lung damage severity (CTSS, opacity, and high opacity, presence of any lesions, GGO and reticulations) and between readouts of acute COVID-19 severity (COVID-19 severity, ICU stay, hospital stay, anti-infective treatment during acute infection) (__Supplementary Figure S17A__). 
In particular, CTSS and opacity (correlation coefficient, Spearman's $\rho$ = `r corr_lft$test$ct %>% filter(variable1 == 'CTSS', variable2 == 'opacity_percent') %>% .$rho %>% signif(2)`), CTSS and high opacity ($\rho$ = `r corr_lft$test$ct %>% filter(variable1 == 'CTSS', variable2 == 'high_opacity_percent') %>% .$rho %>% signif(2)`), and opacity and high opacity (`r corr_lft$test$ct %>% filter(variable1 == 'opacity_percent', variable2 == 'high_opacity_percent') %>% .$rho %>% signif(2)`) were strongly inter-correlated (__Supplementary Figure S3B__, __Supplementary Table S11__). 
CTSS values were significantly higher in observations with than without 
GGO (difference of medians: `r blast(comp_ct$analysis_tbl, GGO_finding) %>% map(~.x$CTSS) %>% map_dbl(median) %>% diff` points, `r comp_ct$test %>% filter(variable == 'CTSS', split_factor == 'GGO_finding') %>% .$significance`, effect size: `r comp_ct$test %>% filter(variable == 'CTSS', split_factor == 'GGO_finding') %>% .$eff_size`) and 
reticulation (difference of medians: `r blast(comp_ct$analysis_tbl, reticulation_finding) %>% map(~.x$CTSS) %>% map_dbl(median) %>% diff` points, `r comp_ct$test %>% filter(variable == 'CTSS', split_factor == 'reticulation_finding') %>% .$significance`, effect size: `r comp_ct$test %>% filter(variable == 'CTSS', split_factor == 'reticulation_finding') %>% .$eff_size`). 
Analogically, significantly higher opacity was detected in observations with than without 
GGO (difference of medians: `r blast(comp_ct$analysis_tbl, GGO_finding) %>% map(~.x$opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`% of lung, `r comp_ct$test %>% filter(variable == 'opacity_percent', split_factor == 'GGO_finding') %>% .$significance`, effect size: `r comp_ct$test %>% filter(variable == 'opacity_percent', split_factor == 'GGO_finding') %>% .$eff_size`) and 
reticulation (difference of medians: `r blast(comp_ct$analysis_tbl, reticulation_finding) %>% map(~.x$opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`% of lung, `r comp_ct$test %>% filter(variable == 'opacity_percent', split_factor == 'reticulation_finding') %>% .$significance`, effect size: `r comp_ct$test %>% filter(variable == 'opacity_percent', split_factor == 'reticulation_finding') %>% .$eff_size`). 
Effect sizes of these differences were large (__Supplementary Figure S17B__, __Supplementary Tables S12__).
Collectively, the high co-linearity of the most important explanatory variables points suggests their redundancy. 
Tackling this redundancy e.g. by regularization may control the over-fitting and improve accuracy of predictions of DLCO and reduced DLCO by machine learning algorithms.

@@

Human-determined CTSS (difference of medians: `r blast(lft_uniboot$analysis_tbl, DLCO_reduced) %>% map(~.x$CTSS) %>% map_dbl(median) %>% diff` points, `r filter(lft_uniboot$test, split_factor == 'DLCO_reduced', variable == 'CTSS')$significance`, effect size: `r filter(lft_uniboot$test, split_factor == 'DLCO_reduced', variable == 'CTSS')$eff_size`), 
AI-determined opacity (difference of medians: `r blast(lft_uniboot$analysis_tbl, DLCO_reduced) %>% map(~.x$opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`% of the lung, `r filter(lft_uniboot$test, split_factor == 'DLCO_reduced', variable == 'opacity_percent')$significance`, effect size: `r filter(lft_uniboot$test, split_factor == 'DLCO_reduced', variable == 'opacity_percent')$eff_size`) and 
high opacity (difference of medians: `r blast(lft_uniboot$analysis_tbl, DLCO_reduced) %>% map(~.x$high_opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`, `r filter(lft_uniboot$test, split_factor == 'DLCO_reduced', variable == 'high_opacity_percent')$significance`% of the lung, effect size: `r filter(lft_uniboot$test, split_factor == 'DLCO_reduced', variable == 'high_opacity_percent')$eff_size`) 
were significantly higher in observations with DLCO < 80% of the patient's reference than in the remaining data points. 
The effect size of those differences was large (__Figure \@ref(fig:fig-dlco-red-detection)A__). 
AI-measured opacity of the lung, but not CTSS or high opacity, was found to be significantly higher in observations with insufficient 
FVC (difference of medians: `r blast(lft_uniboot$analysis_tbl, FVC_reduced) %>% map(~.x$opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`, `r filter(lft_uniboot$test, split_factor == 'FVC_reduced', variable == 'opacity_percent')$significance`, effect size: `r filter(lft_uniboot$test, split_factor == 'FVC_reduced', variable == 'opacity_percent')$eff_size`) and 
insufficient FEV1 (difference of medians: `r blast(lft_uniboot$analysis_tbl, FEV1_reduced) %>% map(~.x$opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`, `r filter(lft_uniboot$test, split_factor == 'FEV1_reduced', variable == 'opacity_percent')$significance`, effect size: `r filter(lft_uniboot$test, split_factor == 'FEV1_reduced', variable == 'opacity_percent')$eff_size`); sizes of those effects were, however, small (__Supplementary Table S13__). 
Furthermore, CTSS, opacity and high opacity correlated negatively with moderate effect size with DLCO expressed as percentage of the patient's reference (p < 0.001, Spearman's $\rho$: `r lft_uniboot$cor_test %>% filter(variable1 == 'DLCO_percent') %>% .$rho %>% range %>% signif(2) %>% paste(collapse = ' to ')`). 
Interestingly, the CT readouts of structural lung damage also correlated significantly with FVC (Spearman's $\rho$: `r lft_uniboot$cor_test %>% filter(variable1 == 'FVC_percent') %>% .$rho %>% range %>% signif(2) %>% paste(collapse = ' to ')`) and FEV1 (Spearman's $\rho$: `r lft_uniboot$cor_test %>% filter(variable1 == 'FEV1_percent') %>% .$rho %>% range %>% signif(2) %>% paste(collapse = ' to ')`); effect size of those associations was small (__Supplementary Table S14__).
